Local wind regime induced by giant linear dunes: comparison of ERA5-Land reanalysis with surface measurements

### 2022-07-07 Paper/main.tex
Adding correction from Clément

\begin{abstract}
Emergence and growth of sand dunes results from the dynamic interaction between topography, wind flow and sediment transport. While feedbacks between these variables are well studied at the scale of a single and relatively small dune, the average effect of a periodic large-scale dune pattern on atmospheric flows remains poorly constrained, due to a lack of data in major sand seas. Here, we compare local measurements of surface winds to the predictions of the ERA5-Land climate reanalysis at four locations in Namibia, within and outside the giant linear dune field of the Namib sand sea. In the desert plains to the north of the sand sea, observations and predictions agree well. This is also the case in the interdune areas of the sand sea during the day. During the night, however, an additional wind component aligned with the giant dune orientation is measured, in contrast to the easterly wind predicted by the ERA5-Land reanalysis.
%
We link these discrepancies, with wind deviation and velocity attenuation larger then $50^\circ$ and $60~\%$, to the daily cycle of the turbulent atmospheric boundary layer over a complex topography, and to the associated flow regimes. During the night, a shallow boundary layer induces a flow confinement associated with a strong streamline compression above the giant dunes, resulting in large flow deviations, especially for the the lower winds. During the day, the flow confinement is reduced by a thicker boundary layer and higher wind velocities, and the feedback of the giant dunes on the atmospheric flow is negligible.
%
We finally propose that this mechanism and the resulting wind deflections by the giant dunes could explain the occurrence of smaller-scale secondary dune patterns, elongating along a different orientation compared to the primary dunes between which they develop.

@@

Compared to hills and mountains, aeolian sand dunes offer idealized elevation profiles for the study of atmospheric turbulent flow over topographies, due to their smooth shape, free of canopies. Besides, dunes provide a rather wide range of scales, from decameters to kilometers, and very often come in a fairly regular pattern, which further simplifies the flow structure analysis. Past studies have highlighted two important topographic feedbacks on the wind flow close to the dune/hill surface. First is the effect on wind speed, with documented flow acceleration on upwind slopes \citep{Weaver2011} and deceleration on downwind slopes \citep{Baddock2007}, where the speed-up factor is essentially proportional to the obstacle aspect ratio \citep{Jackson1975}. Importantly, the velocity maximum is typically shifted upwind of the obstacle crest \citep{Jackson1975, Claudin2013}. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998, Kroy2002}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007, Bristow2022}, in wind tunnels \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions at all scales \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}. The second effect, much less studied, is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as well-developed sand dunes \citep{Tsoar1983, Sweet1990, Walker2002, Smith2017} and in particular coastal foredunes (e.g.~ \citet{Hunter1983}, \citet{rasmussen1989}, \citet{Walker2006}, \citet{Walker2009}, \citet{Hesp2015}, \citet{Walker2017}, \citet{deWinter2020}), mountain ranges \citep{Kim2000, Lewis2008, Fernando2019}, and valley topographies \citep{Wiggs2002, Garvey2005}.

Wind measurements over sand dunes has been mainly performed over small bedforms, typically a few meters high (corresponding to several tens of meters long) (e.g.~\citet{Mulligan1988}, \citet{Hesp1989}, \citet{Lancaster1996}, \citet{MckennaNeuman1997}, \citet{Sauermann2003}, \citet{Andreotti2002}, \citet{Walker2002}, \citet{Weaver2011}). For practical reasons, fewer studies performed similar measurements on giant dunes \citep{Havholm1988}, with kilometer-scale wavelengths and heights of tens of meters. However, such large dunes provide a choice configuration for the study of turbulent flows over a complex topography. First, one expects larger wind disturbances for larger obstacles. Secondly, their large size can make them interact with the vertical structure of the atmosphere \citep{Andreotti2009}. Third, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated dunes. Finally, because the morphodynamics of aeolian bedforms is strongly dependent on the local wind regime \citep{Livingstone2019}, one can expect to see the consequences of windflow disturbance by large dunes on neighbouring small dunes \citep{Brookfield1977, Ewing2006}. A similar effect is observed on the properties of impact ripple patterns due to the presence of dunes \citep{Howard1977, Hood2021}.

@@

It is important to discuss the sensitivity of the results with respect to the choice of the hydrodynamic roughnesses (see Online Resource section~\ref{calib_z0}). In fact, the only quantities dependent on this choice are those which involve the amplitude of the velocities: wind shear velocities, Froude number $\mathcal{F}$ and relative velocity modulation $\delta_{u}$. Those associated with wind direction are independent of this choice. Considering the possible range of realistic roughnesses values, the uncertainty on velocities estimated using the law of the wall is at most $30 \%$. A similar maximum uncertainty applies to the Froude number. This uncertainty also propagates to $\delta_{u}$, for which Figure~\ref{Fig14_supp} shows that the choice of roughness has little influence of its temporal variations, even if it can induce a global increase or decrease of its values. As such, the choice of $z_0$ will not qualitatively affect the overall aspect of the regime diagram presented in Figure~\ref{Fig8}b. It may only change the crossover value of $\delta_{u}$ at which the transition between regimes is observed (dashed green lines in that figure). Our conclusions are thus robust with respect to the somewhat arbitrary choice of the hydrodynamic roughnesses in the use of the data from ERA5-Land reanalysis.

>>>>

\begin{abstract}
Emergence and growth of sand dunes results from the dynamic interaction between topography, wind flow and sediment transport. While feedbacks between these variables are well studied at the scale of a single and relatively small dune, the average effect of a periodic large-scale dune pattern on atmospheric flows remains poorly constrained, due to a lack of data in major sand seas. Here, we compare local measurements of surface winds to the predictions of the ERA5-Land climate reanalysis at four locations in Namibia, within and outside the giant linear dune field of the Namib sand sea. In the desert plains to the north of the sand sea, observations and predictions agree well. This is also the case in the interdune areas of the sand sea during the day. During the night, however, an additional wind component aligned with the giant dune orientation is measured, in contrast to the easterly wind predicted by the ERA5-Land reanalysis.
%
For the given dune orientation and measured wind regime, we link
the observed wind deviation (up to 50\textdegree) to the daily cycle of the turbulent
atmospheric boundary layer. During the night, a shallow boundary layer
induces a flow confinement above the giant dunes, resulting in large
flow deviations, especially for the lower easterly winds. During the
day, the feedback of the giant dunes on the atmospheric flow is much weaker
due to the thicker boundary layer and higher wind speeds. Finally, we
propose that the confinement mechanism and the associated wind
deflections induced by giant dunes could explain the development of smaller-scale secondary
dunes, elongating obliquely in the interdune areas of the primary dune
patterns.

@@

Compared to hills and mountains, aeolian sand dunes offer idealized elevation profiles for the study of atmospheric turbulent flow over topographies, due to their smooth shape, free of canopies. Besides, dunes provide a rather wide range of scales, from decameters to kilometers, and very often come in a fairly regular pattern, which further simplifies the flow structure analysis. Past studies have highlighted two important topographic feedbacks on the wind flow close to the dune/hill surface.
%
First is the effect on wind speed, with documented flow acceleration on upwind slopes \citep{Weaver2011} and deceleration on downwind slopes \citep{Baddock2007}, where the speed-up factor is essentially proportional to the obstacle aspect ratio \citep{Jackson1975}. Under multidirectional wind regimes with frequent wind reversals, this speed-up effect induces large differences in the amplitude and orientation of the resultant sediment transport between flat sand beds and the dune crests \citep{Zhang2014, Rozir2019, Geo2021}. In addition, the maximum in wind velocity is typically shifted upwind of the obstacle crest \citep{Jackson1975, Claudin2013}. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998, Kroy2002}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007, Bristow2022}, in wind tunnels \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions at all scales \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}.

The second effect, much less studied, is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as well-developed sand dunes \citep{Tsoar1983, Sweet1990, Walker2002, Smith2017} and in particular coastal foredunes \citep[e.g.][]{Hunter1983, rasmussen1989, Walker2006, Walker2009, Hesp2015, Walker2017, deWinter2020}, mountain ranges \citep{Kim2000, Lewis2008, Fernando2019}, and valley topographies \citep{Wiggs2002, Garvey2005}.

Wind measurements over sand dunes has been mainly performed over small bedforms, typically a few meters high (corresponding to several tens of meters long) \citep[e.g.][]{Mulligan1988, Hesp1989, Lancaster1996, MckennaNeuman1997, Sauermann2003, Andreotti2002, Walker2002, Weaver2011}. For practical reasons, fewer studies performed similar measurements on giant dunes \citep{Havholm1988}, with kilometer-scale wavelengths and heights of tens of meters. However, such large dunes provide a choice configuration for the study of turbulent flows over a complex topography. First, one expects larger wind disturbances for larger obstacles. Secondly, their large size can make them interact with the vertical structure of the atmosphere \citep{Andreotti2009}. Third, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated dunes. Finally, because the morphodynamics of aeolian bedforms is strongly dependent on the local wind regime \citep{Livingstone2019}, one can expect to see the consequences of windflow disturbance by large dunes on neighbouring small dunes \citep{Brookfield1977, Ewing2006}. A similar effect is observed on the properties of impact ripple patterns due to the presence of dunes \citep{Howard1977, Hood2021}.

@@

As the hydrodynamic roughness $z_0$ determine the magnitude of wind shear velocities, Froude
number $\mathcal{F}$ and relative velocity modulation $\delta_{u}$, it is important to discuss
the sensitivity of the results to the $z_0$-values chosen for both the ERA5-Land and the field
data (see Online Resource section~\ref{calib_z0}). Other quantities associated with wind direction are
independent of this choice. Considering the possible range of realistic roughness values, the
uncertainty on velocities estimated using the law of the wall is at most 30~\%. A similar
maximum uncertainty applies to the Froude number. This uncertainty also propagates to
$\delta_{u}$, for which Figure~\ref{Fig14_supp} shows that the choice of roughness has little influence on
its temporal variations,even if it can induce a global increase or decrease of its values. Hence, the choice
of the $z_0$-values will not qualitatively affect the overall aspect of the regime diagram
presented in Figure 8b. It may only change the value of $\delta_u$ for which the transition
between regimes is observed (dashed green lines in Figure~\ref{Fig8}b). Our conclusions are thus robust
with respect to the somewhat arbitrary choice of the hydrodynamic roughness values.
